---
title: "Statistical Analysis of Body Composition & Health Indicators"
author: "Rasha Quadri"
date: "2026-03-22"
output: 
  html_document:
    toc: true
    toc_depth: 2
    theme: readable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# **Introduction**

Understanding how different aspects of body composition relate to one another is important in health and medical analysis.

This project analyzes relationships between key body composition variables, including **body protein content**, **total body fat ratio**, and **total body water**. 

The objectives of this analysis are to:

- Identify relationships between variables
- Build and evaluate regression models
- Assess statistical assumptions
- Interpret results in a meaningful real-world context

Statistical techniques such as **linear regression**, **multiple regression**, and **ANOVA** are used to extract insights from the dataset.




# **Data Preprocessing**


```{r}

medical <- read.csv("medicaldata2.csv", header = TRUE)

head(medical)

```

## **Converting Variables to Factors**

```{r}

medical$Gallstone_Status <- factor(medical$Gallstone_Status)
medical$CAD <- factor(medical$CAD)
medical$Diabetes_Mellitus <- factor(medical$Diabetes_Mellitus)

medical$Comorbidity <- factor(
  medical$Comorbidity,
  ordered = TRUE,
  levels = c(0 , 1, 2, 3)
)

medical$Hepatic_Fat_Accumulation <- factor(
  medical$Hepatic_Fat_Accumulation,
  ordered = TRUE,
  levels = c(0, 1, 2, 3)
)

medical$Age_Class <- factor(
  medical$Age_Class,
  ordered = TRUE,
  levels = c("Younger", "Lower Middle", "Upper Middle", "Elder")
)


```

Categorical variables were converted into factors to ensure proper handling in statistical models.



# **Simple Linear Regression**

## **Relationship: Body Protein vs Body Fat**

```{r}

plot(
  medical$Body_Protein_Content,
  medical$Total_Body_Fat_Ratio,
  xlab = "Body Protein Content",
  ylab = "Total Body Fat Ratio",
  main = "Body Fat Ratio vs Body Protein Content"
)

```

```{r}
model1 <- lm(Total_Body_Fat_Ratio ~ Body_Protein_Content, data = medical)
summary(model1)

```


## **Model Diagnostics**

```{r}

plot(model1$fitted.values, model1$residuals)
abline(h=0,col="red")

qqnorm(model1$residuals)
qqline(model1$residuals, col="red")

shapiro.test(model1$residuals)

```

```{r}

cor(medical$Total_Body_Fat_Ratio, medical$Body_Protein_Content)

```

## **Interpretation**

There is a clear **negative linear relationship** between body protein content and total body fat ratio.

- The model explains about **49% of the variation** (R^2^ = 0.49)
- The predictor is **statistically significant** (p < 0.05)
- Correlation confirms a **strong negative association**

Residual diagnostics show no major violations of assumptions.

**Conclusion**: Higher body protein content is associated with lower body fat.



# **Multiple Linear Regression**

## **Predicting Total Body Water**

```{r}

model2 <- lm(Total_Body_Water ~ Height + Weight + Body_Protein_Content, data = medical)
summary(model2)

```


## **Diagnostics**

```{r}

plot(model2$fitted.values, model2$residuals)
abline(h=0, col="red")

qqnorm(model2$residuals)
qqline(model2$residuals)

shapiro.test(model2$residuals)

```

```{r}

library(lmtest)
dwtest(model2)

```

## **Interpretation**

The model explains approximately **78% of the variation** in total body water.

- **Height and Weight** are statistically significant predictors
- **Body Protein Content is not significant**
- No evidence of autocorrelation

**Conclusion**: Body water is primarily driven by physical characteristics (height and weight), not protein content.


# **ANOVA: Age and Body Fat**

## **Research Question**

Does body fat differ across age groups?

```{r}

library(car)

anova_model <- aov(Total_Body_Fat_Ratio ~ Age_Class, data = medical)
summary(anova_model)

```


## **Assumption Checks**

```{r}

qqnorm(residuals(anova_model))
qqline(residuals(anova_model))

shapiro.test(residuals(anova_model))

leveneTest(Total_Body_Fat_Ratio ~ Age_Class, data = medical)

```

## **Visualization**

```{r}

library(ggplot2)

ggplot(medical, aes(x=Age_Class, y=Total_Body_Fat_Ratio))+
  geom_boxplot(fill="lightblue")+
  theme_minimal()

```

## **Interpretation**

The ANOVA test is **statistically significant (p < 0.05)**, indicating that body fat differs across age groups.

Post-hoc comparisons show that **middle age groups tend to have higher body fat levels** compared to younger individuals.

**Conclusion**: Age plays a meaningful role in body fat distribution.


# **Two-Way ANOVA: Triglyceride Levels**

## **Research Question**

Do diabetes and hepatic fat affect triglyceride levels?

```{r}

model3 <- aov(Triglyceride ~ Diabetes_Mellitus * Hepatic_Fat_Accumulation, data = medical)
summary(model3)

```
## **Assumptions**

```{r}

shapiro.test(residuals(model3))

qqnorm(residuals(model3))
qqline(residuals(model3))

leveneTest(Triglyceride ~ Diabetes_Mellitus * Hepatic_Fat_Accumulation, data = medical)

```

## **Visualization**

```{r}

ggplot(medical, aes(x=Diabetes_Mellitus, y=Triglyceride))+
  stat_summary(fun=mean, geom="bar")

ggplot(medical, aes(x=Hepatic_Fat_Accumulation, y=Triglyceride))+
  stat_summary(fun=mean, geom="bar")

```

## **Interpretation**

- The **interaction effect is not significant**, meaning the effect of diabetes does not depend on hepatic fat level

- Both variables have **significant individual effects**

**Conclusion:**

- Diabetes is associated with higher triglyceride levels
- Hepatic fat accumulation also increases triglycerides
- These effects act independently

# **Key Insights**

- Body protein is strongly associated with **lower body fat**
- Height and weight are the **main drivers of body water**
- Age significantly affects **body fat distribution**
- Diabetes and hepatic fat both impact **triglyceride levels**

# **Conclusion**

This project demonstrates how statistical modeling techniques can be used to analyze real-world health data.

By applying regression and ANOVA methods, we identified key predictors, evaluated assumptions, and extracted meaningful insights from the dataset.

These methods provide a strong foundation for data-driven decision making in health and medical contexts.